\section{Where Does it Start?}

There is significant debate about who created the first compiler,in no small
part due to ambiguous nomenclature. The term \textit{software} came into use
sometime between 1959 and 1962, as \citeauthor{the-first-computers-2002} note:

\begin{quotation}
	[Expressions] such as "hardware", "software", "machine language",
	"compiler", "architecture" and the like... were unknown in 1950.
	They only arrived a decade later, but the underlying concepts were
	quite familiar to us.
	\cite{the-first-computers-2002}
\end{quotation}

This Honeywell advertisement \textit{A Few Quick Facts on Software} sought to
clarify these terms as well:

\begin{quotation}
	Software is a new and important addition to the jargon of computer
	users and builders. It refers to the automatic programming aids that
	simplify the task of telling the computer 'hardware' how to do its
	job.
	\cite[ch.5]{new-history-of-modern-computing}
\end{quotation}

At this time, hardware was the only piece that mattered to customers. Software
was an afterthought, if a thought at all. The instruction set of the machine was
important, because that was the user's interface to the machine. It should come
as no surprise, then, that the origins of our modern understanding of the term
\textit{compiler} are similarly murky, especially considering the fact that
\textit{compiler} already carried meaning in English, and was repurposed for
computing. John Backus even pointed out how the ambiguity around the term
\textit{compiler} makes computing history circa 1950s especially difficult to
untangle \cite{Backus_1980_Programming_in_America_in_1950s}:

\begin{quotation}
	There is an obstacle to understanding, now, developments in
	programming in the early 1950s. There was a rapid change in the
	meaning of some important terms during the 1950s. We tend to assume
	that the modern meaning of a word is the same one it had in an early
	paper, but this is sometimes not the case. Let me illustrate this
	point with examples concerning the word "compiler."
\end{quotation}

\bigskip
Once could argue that any of these efforts constituted the first compiler:
\begin{itemize}
	\item Konrad Zuse's run-programs for the Z4 at the ETH Zurich in 1950
	\item Grace Hopper's A-0 and A-1 compilers for the UNIVAC I at
	      Remington Rand in 1951
	\item Laning and Zierler's algebraic compiler for the Whirlwind at MIT in 1950
	\item John Backus's \FTNI{} compiler at IBM
\end{itemize}

I attempt to discuss these in order, however their efforts overlap
significantly in time. I try to tell their stories as a whole, though each story
contains references to the others;if you find yourself confused by names
introduced out of context, please finish the chapter or search for the content
within this chapter before giving up.

\section{Konrad Zuse and the Z4}

The historians Paul Ceruzzi and Raul Rojas have written extensively about
Zuse and his work on the Z3, Z4, and the special-purpose S1 and S2 machines--
the reader is encouraged to consult their works for a more exhaustive account
\cite{ceruzzi_early_computers_of_zuse_1981}
\cite{rojas_arch_of_zuses_machines_2000},
as well as Konrad's son's account of his father's life\cite{zuse2009life}.

We will meet Konrad Zuse, a German civil engineer, during World War II when
he began work on the Z4.
Funded partially by his family and partially by the Nazi government, his
prior works demonstrated significant creativity and ingenuity, and they were
leveraged to build precursors to modern cruise missiles and guided bombs. Most of
Zuse's machines prior to the Z4 were destroyed during the war;
he rescued his Z4 by moving it himself to a remote alpine village
called Hinterstein\cite{Knuth_TrabbPardo_1976_Early_Development}.
While he was funded by the Third Reich at some point, he was primarily \textit{denied}
funding by the government. In a lecture in New Mexico in 1976, he claimed
that his computers were never used to carry out their \textit{final solution}.

At this time, computers were transitioning from \textit{sequenced calculators}
(capable only of executing straight-line arithmetic calculations)
to "true" computers capable of control flow.
The Z3 was unique in its ability to perform conditional jumps, qualifying it
as a "true" computer, more advanced than the sequenced calculators.
It also had floating-point arithmetic units, which were missing from other
early computers like Commander Howard Aiken's Mark I at Harvard and IBM's 703.

After the war, he began looking for a market for some of his wartime inventions.
His machines were unique, containing floating-point arithmetic units and a
bespoke memory system.
From 1943 to 1945, he worked on the \textit{Plankalkül} (for \textit{Plan Calculus}),
an algebraic programming language and possibly the first high-level programming language,
though it was not implemented until 1975 in Joachim Hohmann's PhD dissertation
\citetitlecite{hohmann1979plankalkul}.

Towards the end of the war, Zuse and his pregnant wife fled Berlin.
Thanks to some confusion about the purpose of Zuse's machine, he was able to leverage
transport vehicles from the German army.
At the time, the German military was working on a new generation of rocket, which
went by the name \textit{V2}. At the time, Konrad named his machines V1, V2, and so on,
where V stood for \textit{Versuchsmodell}, or \textit{trial models}.
Thus the German army believed his machine was somehow related to the rocket program,
and he was able to commandeer transport vehicles from the army.

He was originally to take his machine to the underground factories in Nordhausen
where the V2 rocket was under production. After being shocked by \textit{twenty thousand}
concentration camp workers toiling in miles of underground tunnels, he
decided to take his wife and computer as far away as possible.
Thus they fled to Hinterstein, Germany and then Hopferau, Bavaria, to
hopefully restore the machine.

It is unclear how Professor Stiefel learned of Zuse's Z4, but it was in Hopferau
where he met Konrad Zuse and first saw the Z4 in action.

\iffalse
	In a turn of events Konrad Zuse was made
	aware of Aiken's Mark I through his daughter
	\cite{howard_aiken_and_the_dawn_of_the_computer_age_2000}:

	\begin{quotation}
		Konrad Zuse told me an amusing anecdote about how he first
		encountered the work of Aiken. The occasion of our conversation was
		a luncheon in Zuse's honor, hosted by Ralph Gomory at the Watson
		Research Laboratory of IBM before a lecture given by Zuse to the
		staff of the lab. When Zuse learned that I was gathering materials
		for a book on Aiken, he told me that he had come across Aiken and
		Mark I in an indirect manner, through the daughter of his
		bookkeeper. She was working for the German Secret
		Service (Geheimdienst) and knew through her father of Zuse's work on
		a large scale calculator. According to Zuse,the young woman never
		learned any details about his machine, which was shrouded in
		war-time secrecy. But she knew enough about Zuse's machine to
		recognize that the material filed in a certain drawer related to
		advice that seemed somewhat like Zuse's. She reported this event
		to her father, giving the file number of the drawer, and the father
		at once informed Zuse of her discovery. Zuse, of course, could not
		go to the Secret Service and ask for the document since that would
		give away the illegal source of his information. Zuse was well
		connected, however, and was able to send two of his assistants to
		the Secret Service, armed with an official demand for information
		from the Air Ministry, requesting any information that might be in
		the files concerning a device or machine in any way similar
		to Zuse's. Zuse's assistants were at first informed that no such
		material existed in the files, but they persisted and eventually got
		to the right drawer. There they found a newspaper clipping (most
		likely from a Swiss newspaper), containing a picture of Mark I and a
		brief description about Aiken and the new machine. But there was not
		enough technical information to enable Zuse to learn the machine's
		architecture.
	\end{quotation}
\fi

\subsection{The ETH's Acquisition of the Z4}

There were several early efforts to create programs that produced punch
cards which contained machine code instructions, which could then be fed back
into the machine as input punchcards. The programs produced by these early
compilers were called \textit{run-programs}, and the process of using them was
called \textit{automatic programming}, a term later coined by Grace Hopper. The
first of these programs was run on a machine called the Z4, designed by Konrad
Zuse in Germany.

Professor Eduard Stiefel, shortly after establishing the
Institute of Applied Mathematics to study numerical analysis at the Swiss
Federal Institute of Technology (ETH) in Zurich, began searching for a computer
for the institute. He learned of the computing advancements in the United
States, Great Britain, and Germany,but no machines were readily available at
the time. He sent his assistants Heinz Rutishauser and Ambros P. Speiser to the
US to study the latest developments in computing; they spent most of 1949
with Howard Aiken at Harvard and John von Neumann at Princeton.

\begin{quotation}
	Before we returned, that is, in the middle of 1949, Stiefel was informed
	about the existence of Konrad Zuse's Z4. At that time Zuse was living in
	Hopferau, a German village near the Swiss border. Stiefel was told that the
	machine might be for sale. He visited Zuse, inspected the device, and reviewed
	the specifications. Despite the fact that the Z4 was only barely
	operational, he
	decided that the idea of transferring it to Zurich should by all means be
	considered. Stiefel wrote a letter to Rutishauser and me (we were at Harvard at the time),
	describing the situation and asking us to get Aiken's opinion.
	Aiken's reply was very critical - the future belonged to electronics
	and, rather
	than spending time on a relay calculator, we should now concentrate our efforts
	on building a computer of our own.
	\cite{konrad-zuses-z4-2000}
\end{quotation}

The Z4 was a bespoke machine with unique components; the computational logic
was wired together with telephone relays and the memory was entirely mechanical.

\begin{quotation}
	The Z4 could be used as a kind of manually triggered calculator: the
	operator could enter decimal numbers through the decimal keyboard, these
	were transformed into the floating point representation of the Z4, and were
	loaded to the CPU registers, first to OR-I, then to ORII. Then, it was possible
	to start an operation using the "operations keyboard" (an addition, for
	example). The result was held in OR-I and the user could continue loading
	numbers and computing. The result in OR-I could be made visible in decimal
	notation by transferring it to a decimal lamp array (at the push of a button).
	It could also be printed using an electric typewriter.
	\cite{architecture-of-konrad-zuses-z4-computer-2021}
\end{quotation}

It notably featured instructions for conditional branching and subroutine
calling, which both proved essential for the compiler development that would
follow at the ETH. Stiefel was undeterred by Aiken's criticism, and convinced
the ETH to purchase the Z4. In 1950, Heinz Rutishauser at Switzerland's ETH
obtains a Z4.

\begin{quotation}
	We also made some hardware changes. Rutishauser, who was exceptionally
	creative, devised a way of letting the Z4 run as a compiler, a mode
	of operation
	which Zuse had never intended. For this purpose, the necessary
	instructions were
	interpreted as numbers and stored in the memory. Then, a compiler
	program calculated the program and punched it out on a tape. All this required
	certain hardware changes. Rutishauser compiled a program with as many as 4000
	instructions. Zuse was quite impressed when we showed him this achievement.
	\cite{konrad-zuses-z4-2000}
\end{quotation}

Thus were the first run-programs produced. This is what we will
consider \textbf{the first compiler}, though it was not called that
at the time. Shortly after Stiefel's assistants' stints in the US and
correspondence with Aiken,one of Aiken's engineers would find
considerably more success exploring related ideas.
