\section{Low-Level Virtual Machine}

Key innovation is that LLVM is a collection of compiler \textit{libraries} that
have thin programs wrapping them.Other open-source compilers like GCC tend to be
monolithic programs, which can be harder to compose.LLVM contains a collection
of sub-projects that can be used independently as tools \textit{or libraries};
key examples being the LLVM sub project of LLVM itself (which contains LLVM \GLS{ir},
the intermediate representation, the optimizer, interpreter and code
generator) and Clang, a C/C++/Objective-C compiler front-end for LLVM.Any other
compiler project could emit LLVM IR and fully leverage the LLVM project after
the front-end.LLVM's intermediate representation has become the lingua-franca
of the compiler world.

\begin{quotation}
	From its beginning in December 2000,
	LLVM was designed as a set of reusable libraries with well-defined interfaces
		[LA04]. At the time, open source programming language implementations were
	designed as special-purpose tools which usually had monolithic executables. For
	example, it was very difficult to reuse the parser from a static compiler (e.g.,
	GCC) for doing static analysis or refactoring. While scripting languages often
	provided a way to embed the ir, runtime, and interpreter into larger applications,
	this runtime was a single monolithic lump of code that was included or excluded.
	There was no way to reuse pieces, and very little sharing across language
	implementation projects.\cite{aosa_vol1}
\end{quotation}

\begin{quotation}
	The name "LLVM" was once an acronym, but is now just a brand for the
	umbrella project. While LLVM provides some unique capabilities, and is known for
	some of its great tools (e.g., the Clang compiler2, a C/C++/Objective-C compiler
	which provides a number of benefits over the GCC compiler), the main thing that
	setsLLVM apart from other compilers is its internal
	architecture.\cite[LLVM]{aosa_vol1}
\end{quotation}

Lattner published his thesis
on LLVM in 2002, and joined Apple in
2005.
