\chapter{Dawn, 1940-1960}

\section{Where Does it Start?}

There is significant debate about who created the first compiler,
in no small part due to ambiguous nomenclature.
The term \textit{software} came into use sometime between 1959 and 1962,
as \citeauthor{the-first-computers-2002} note:

\begin{quotation}
[E]xpressions such as "hardware", "software", "machine language", "compiler",
"architecture" and the like... were unknown in 1950. They only arrived a decade
later, but the underlying concepts were quite familiar to us.
\cite{the-first-computers-2002}
\end{quotation}

This Honeywell advertisement \textit{A Few Quick Facts on Software} sought to clarify these terms as well:

\begin{quotation}
Software is a new and important addition to the jargon of computer users and builders.
It refers to the automatic programming aids that simplify the task of telling the computer 'hardware' how to do its job....
% Generally there are three basic categories of software:
% 1) Assembly Systems, 2) Compiler Systems, and 3) Operating Systems.\cite[ch.5]{new-history-of-modern-computing}
\end{quotation}

At this time, hardware was the only piece that mattered to customers.
Software was an afterthought, if a thought at all.
The instruction set of the machine was important, because that was the
user's interface to the machine.
It should come as no surprise, then, that the origins of our modern understanding of the
term \textit{compiler} are similarly murky, especially considering the fact that
\textit{compiler} already carried meaning in English, and was repurposed for computing.

\bigskip

Once could argue that any of these efforts consituted the first compiler:
\begin{itemize}
    \item Konrad Zuse's run-programs for the Z4 at the ETH Zurich in 1950
    \item Laning and Zierler's compiler for the Whirlwind at MIT in 1950
    \item Grace Hopper's A-0 and A-1 compilers for the UNIVAC I at Remington Rand in 1951
\end{itemize}

% We will discuss these in order \todo{and then COBOL and Backus+\FTN}.

\section{Development of the Z4}

Konrad Zuse, a German civil engineer, began work on the Z4 during World War II.
Funded partially by his family and partially by the Nazi government, his prior works
demonstrated significant creativity and ingenuity, and they were leveraged to build
precursors to modern cruise missiles and guided bombs.
Most of Zuse's machines prior to the Z4 were destroyed during the war.

\todo{
Zuse's Z4 was a strange machine with bespoke memory and instruction set.
This affected how the compilers for it were designed.
}

\section{The ETH's Aquisition of the Z4}

There were several early efforts to create programs that produced punch cards
which contained machine code instructions, which could then be fed back into the machine
as input punchcards.
The programs produced by these early compilers were called \textit{run-programs},
and the process of using them was called \textit{automatic programming}, a term later
coined by Grace Hopper.
The first of these programs was run on a machine called the Z4, designed by Konrad Zuse in Germany.

Professor Eduard Stiefel, shortly after establishing the Institute of Applied Mathematics
to study numerical analysis at the Swiss Federal Institute of Technology (ETH) in Zurich,
began searching for a computer for the institute.
He learned of the computing advancements in the United States, Great Britain, and Germany,
but no machines were readily available at the time.
He sent his assistants Heinz Rutishauser and Ambros P. Speiser to the US to study
the latest developments in computing; they spent most of 1949 with
Howard Aiken at Harvard and John von Neumann at Princeton.

\begin{quotation}
Before we returned, that is, in the middle of 1949, Stiefel was informed about the existence of Konrad Zuse's
Z4. At that time Zuse was living in Hopferau, a German village near the Swiss border. Stiefel was told that
the machine might be for sale. He visited Zuse, inspected the device, and reviewed the specifications.
Despite the fact that the Z4 was only barely operational, he decided that the idea of transferring it to Zurich
should by all means be considered. Stiefel wrote a letter to Rutishauser and me (we were at Harvard at the
time), describing the situation and asking us to get Aiken's opinion. Aiken's reply was very critical - the
future belonged to electronics and, rather than spending time on a relay calculator, we should now
concentrate our efforts on building a computer of our own.\cite{konrad-zuses-z4-2000}
\end{quotation}

The Z4 was a bespoke machine with unique components; the computational logic was wired together with
telephone relays and the memory was entirely mechanical.

\begin{quotation}
The Z4 could be used as a kind of manually triggered calculator: the operator 
could enter decimal numbers through the decimal keyboard, these were 
transformed into the floatingpoint representation of the Z4, and were loaded to 
the CPU registers, first to OR-I, then to ORII. Then, it was possible to start 
an operation using the "operations keyboard" (an addition, for example). The 
result was held in OR-I and the user could continue loading numbers and 
computing. The result in OR-I could be made visible in decimal notation by 
transferring it to a decimal lamp array (at the push of a button). It could 
also be printed using an electric typewriter. \cite{architecture-of-konrad-zuses-z4-computer-2021}
\end{quotation}

It notably featured instructions for conditional branching and subroutine calling, which
both proved essential for the compiler development that would follow at the ETH.
Stiefel was undeterred by Aiken's criticism, and convinced the ETH to purchase the Z4.
In 1950, Heinz Rutishauser at Switzerland's ETH obtains a Z4.

\begin{quotation}
We also made some hardware changes. Rutishauser, who was exceptionally creative, devised a way of
letting the Z4 run as a compiler, a mode of operation which Zuse had never intended. For this purpose, the
necessary instructions were interpreted as numbers and stored in the memory. Then, a compiler program
calculated the program and punched it out on a tape. All this required certain hardware changes. Rutishauser
compiled a program with as many as 4000 instructions. Zuse was quite impressed when we showed him this
achievement.\cite{konrad-zuses-z4-2000}
\end{quotation}

Thus were the first run-programs produced.
This is what we will consider \textbf{the first compiler}, though it was not called that at the time.
Shortly after Stiefel's assistants' stints in the US and correspondence with Aiken,
one of Aiken's engineers would find considerably more success exploring related ideas.

\section{Grace Hopper}

While Grace Hopper may not have been the first to create a program that
punched out another program as its output, she pioneered the field of
compilation to the extent that many consider her the inventor of compilers.
Her innovations were also more readily adopted than those at the ETH.
Consider Figure \ref{fig:dawn-timeline}, and the pace of development in Hopper's time
compared to the years prior.
Note that while we have ample data on \textit{how} Hopper's compiler worked and how
she and her team developed it, the intuition behind those developments is foggy at best.
We have recollections from Hopper and her contemporaries, but only from long after the fact.
It was not understood at the time how important her work was, so we have
only to speculate and piece together oral histories.

Originally a mathematics professor at Vassar College, Hopper obtained waivers for her age and weight and
joined the U.S. Navy in 1943, eventually graduating first in her class from Midshipmen's School.
She was assigned, somewhat unexpectedly, to Commander Howard Aiken's Harvard Computation Laboratory in 1944
as the third programmer of the Automatic Sequence Controlled Calculator (Mark I), 
the world's first operational computer.
Although it was significantly slower than the ENIAC, it was \textit{programmable};
the ENIAC had to be physically rewired to change its program.

Aiken built it in collaboration with IBM, though it is unclear how much either side contributed
in its development.
The proportion of credit given to either party would be disputed in numerous documents and press releases
in the following years, including the
\textit{Manual of Operation for the Automatic Sequence Controlled Calculator},
the technical manual for the Mark I.
In the fall of 1944, Aiken decided that his team needed to produce a book documenting the technical
developments at the Harvard Computation Laboratory and how the Mark I was intended to be used.
For this task, he chose Hopper; though she protested, her reputation as a clear and
thoughtful communicator and writer had already earned her the job.
Hopper was known for her writing ability, and it was a point of emphasis in her teaching career.
She would assign her mathematics students onerous writing tasks to emphasize that
"it was no use trying to learn math unless they could communicate with other people."
\cite[interview on 5 July, 1972]{grace_hopper_and_the_invention_of_the_information_age_2009}

Aiken and Hopper both understood that, for the Mark I to be a success, it would be used and
understood by a large and diverse audience, and for that to happen, they needed a detailed,
compelling, and accessible manual.

% \todo{
% ENIAC, could not be programmed (had to be rewired to be programmed)
% but was real fast. Aiken's Mark I.
% Fall 1944, Aiken wants to record technical developments at Harvard Computation Laboratory (as book),
% chose Hopper, to her disagreement. Finished 561 page manuscript, published 1946.
% She believed she was chosen for "clear, fluid prose."
% her time as a math teacher made her good at writing/explaining things.
% would force students to write essays on math topics and would grade harshely.
% "it was no use trying to learn math unless they could communicate with other people."
% Aiken and Hopper knew a large/diverse audience would need to understand it for it to
% be successful; high value on good writing+communication.
% Aiken's strict heirarchy+meritocracy allowed her to compete with the men.
% IBM tried to steal all the credit for the Mark I from Aiken.
% aiken distaste for being assigned woman officer, but meritocracy.
% }

\section{Hopper and the Mark I's Manual}

Here we drift into some general computing history, mostly because it was so 
formative for Hopper, who was in turn so formative in the development of 
compilers.
Her manual for the Mark I began with a detailed and dramatic retelling of
computing history, opening with the following quote from Charles Babbage and
culminating with the Mark I
\cite{annals_of_the_computation_laboratory_of_harvard_university_1946}:

\begin{quotation}
If, unwarned by my example, any man shall undertake and shall succeed in 
really constructing an engine embodying in itself the whole of the executive 
department of mathematical analysis upon different principles or by simpler 
mechanical means, I have no fear of leaving my reputation in his charge, for he 
alone will be fully able to appreciate the nature of my efforts and the value 
of their results.
\end{quotation}

Her history went on to cover:

\begin{itemize}
    \item Blaise Pascal's counting machines, "foundation on which nearly all mechanical calculating machines since have been constructed."
    \item Leibniz; stepped wheels system for mul/divs.
    \item Charles Babbage; most significant part of the manual dedicated to him.
          Difference engine, idea for computing machine. Invented punchcard system to feed in information, made after textile looms.
          G H emphasized the machine would take 2 decks of cards, one for data, one for instructions (not von neumann).
    \item Ada King, Countess of Lovelace; series of essays on Babbage's machine.
          described possibly the first computer program. This could never run and would
          have to wait for the Mark I before the dream could come true.
    \item Aiken's Mark I
\end{itemize}

At one point, all new hires into Aiken's lab were required to read
Charles Babbage's autobiography.
Hopper was first exposed to Ada King in this text:
"she wrote the first loop. I will never forget; none of us ever will."
\todo{Coworkers cast Aiken as Babbage, Hopper as Ada King.}

\todo{Commander Edmund Berkeley, 1946, detailed report about how bad the
conditions were in the lab.}
\begin{quotation}
In his report, Berkeley systematically detailed the unfavorable conditions at 
the Computation Laboratory, including the length of the work day and the 
isolation of the staff from similar projects at MIT and the University of 
Pennsylvania. He named eleven talented people who had left or been dismissed by 
Aiken between August 1945 and May 1946, noting that all were "very bitter over 
the conditions on the project." The root of the problem, according to Berkeley, 
was that "in the Computation Laboratory there is no provision for appealing any 
decision or ruling whatsoever made by the project manager." He was amazed that 
no one at Harvard and no one in the Navy seemed to have jurisdiction over the 
rogue director, so that Aiken was able to rule with near absolute authority.
\end{quotation}

Aiken ran a rigidly heirarchical and meritocratic lab, which allowed Hopper
to produce quality work and placed her on more equal footing with her
male coworkers; Aiken openly disliked being assigned a female officer,
but Hopper's competence outweighed any such sentiments.
Her competence did not, however, shield her from the pressures of the environment.
She leveraged her computing knowledge to assist the war effort,
which included the bombing of Nagasaki.
The stresses of the war effort and Aiken's overbearing management drove her to
substance abuse in this time period.

\section{Postwar Collaboration}

Aiken started a symposium on large scale digital computing machinery in 1947 to foster
collaboration that was not possible during the war.

\todo{
1947 Aiken grew in stature. Could relocate military staff to Harvard, which he did with Hopper and two others.
1944-1947 Presper Eckert, John Mauchly, and female operators proved valuable bc of how well they could use computers.
this generated a wave of new projects:
BINAC+UNIVAC at ECC, Mark II and Mark III at Harvard, SSEC at IBM, Whirlwind at MIT,
MANIAC at Inst for Advanced Study, ERA (engineering research associates).
everyone was isolated during the war, now aiken wanted to start sharing ideas.
Symposium on Large Scale Digital Calculating Machinery, June 1947.
}
\begin{quotation}
We'd all been isolated during the war, you see, classified contracts and 
everything under the sun. It was time to get together and exchange information 
on the state of the art, so that we could all go on from there.
\end{quotation}

\todo{connect her discontent at harvard to her leaving.}

After a brief stint of unemployment, Hopper joined a startup called the Eckert-Mauchly Computer Corporation (EMCC)
where she found a more congenial environment to continue her work on compilers.

\begin{quotation}
According to her friend and former Harvard colleague Edmund Berkeley, Hopper 
turned to alcohol during this period as a way to deal with the compounding 
pressures at the Harvard Computation Laboratory. She had dedicated herself 
fully to the overwhelming task of bringing Howard Aikens machines to life.  She 
used the machines to solve critical military problems, including one that 
resulted in an explosion over Nagasaki.  As the psychological strains became 
increasingly pronounced, alcohol seemed to serve as an effective outlet, 
freeing Hopper to express emotions and to temporarily forget obstacles real and 
imagined.  According to Berkeley, the expiration of Hoppers Harvard research 
contract was the best thing that could have happened to her, although in the 
short term unemployment added to the stress.  During the last week of May 1949, 
the 43-year-old programmer packed up her belongings, headed to Philadelphia, 
and bet her future on two younger men who believed they could create the first 
commercial computer 
company.\cite{grace_hopper_and_the_invention_of_the_information_age_2009}
\end{quotation}

Hopper's contract at Harvard expired in May 1949, and seeking a new path, she 
moved to Philadelphia to join the Eckert-Mauchly Computer Corporation (EMCC) as 
a senior mathematician

\section{Hopper at EMCC}

The company Hopper joined was one of the earliest pure computer-focused ventures,
founded by J. Presper Eckert Jr. and John Mauchly
(designers of the ENIAC, or the Electronic Numerical Integrator and Computer).
This startup environment contrasted sharply with the academic rigor of Harvard and the industrial scale of IBM.
There she found an open-minded and welcoming environment to develop her ideas;
Mauchly, who was to become Hopper's boss, was characterized as 
"very broadminded, very gentle, very alive, very interested, very forward looking,"\cite{grace_hopper_and_the_invention_of_the_information_age_2009}
creating a tolerant, flexible company atmosphere in contrast to the pressure she experienced at Harvard.
A majority of their programming staff consisted of mathematically inclined women
who had served as ENIAC operators at the Moore School.

When Hopper arrived in 1949, EMCC had two major projects underway:
the BINAC (Binary Automatic Computer), which was close to completion,
and the UNIVAC I (Universal Automatic Computer), which would be running within a year.
The work environment and upcoming UNIVAC project excited Hopper and enticed her to join the company
after walking out of an interview with IBM.

While the organization was grounds for fruitful and innovative research and development team,
EMCC was under financial strain; they depended on partial payments for UNIVAC-I orders to stay afloat.
The unexpected death of EMCC's chairman Henry Straus forced Eckert and Mauchly to seek a buyer,
which they found in Remington Rand, a typewriter and office equipment manufacturer.

% \section{Hopper and the A-series}

Hopper and her team at Remington Rand developed three "compilers"
in rapid succession, the A-0, A-1, and A-2, for the UNIVAC I.
I quote "compilers" because the A-0 and A-1 were not compilers in the modern sense.
Her work was grounded in intellectual openness, collaboration, and accessability;
she pioneered the debuggability of programming languages, compiler error reporting,
and new ways to share code and collaborate, for example.

\begin{quotation}
Hopper's recollections point to motivations
ranging from an altruistic desire to allow "plain, ordinary people"
to program to dealing with her own laziness. Naturally one must
be skeptical of such claims, for they were made years after the
fact. In 1951 it was difficult for even a visionary like Hopper to
imagine the eventual ubiquity of computer technology, and one
can be pretty confident that Hopper was not a lazy person.\cite{grace_hopper_and_the_invention_of_the_information_age_2009}
\end{quotation}

She began work on the A-0 in October 1951 in her spare time in order to address
the mounting crisis facing Remington Rand: they were unable to fully support their
customers, and their sales teams were, to put it kindly, incompetent with respect
to their product, and they supported their customers as well as one might expect.

\begin{quotation}
Another Hopper programmer, Adele Mildred Koss, was
assigned to Commonwealth Edison when the utility approached
the Chicago sales office concerning a potential purchase of a
UNIVAC for billing and payroll. At the time, Koss was 7 months
pregnant and working part time. Since her pregnancy precluded
travel, Commonwealth Edison management was forced to come
to Philadelphia in order to discuss their billing needs. In the end,
the utility did not buy a UNIVAC, but instead purchased an IBM
701 when it became available. Koss recalled: "I remember Grace
Hopper's memo to management saying 'This is a multi-million
dollar client and you are not treating them like one. You have
only assigned a part time programmer to work with.'"
\cite[Adele Mildred Koss, interviewed by Kathy Kleiman]{grace_hopper_and_the_invention_of_the_information_age_2009}
\end{quotation}

Shortly after this fiasco, Hopper's team was tasked with supporting UNIVAC I customers
at the US Census Bureau, a task she thought no one in the company was prepared for.

\begin{quotation}
Inspired by Holberton's Sort-Merge Generator, Hopper conceived the idea of writing a 
program to create a program, or in modern day terms, building a compiler. 
The idea was to get commonly used subroutines automatically inserted into another program based on calculated offsets. 
Most people at the time considered this impossible. As Hopper recalled:\cite{women_in_computing_history_2002}
\end{quotation}

\begin{quotation}
The Establishment promptly told us, at least they told me, quite frequently that a
computer could not write a program; it was totally impossible; that all computers
could do was arithmetic, and that it couldn't write programs.\cite{hopl_keynote}
\end{quotation}

\todo{
the A-2 was a front-end translator to the A-0, more like a complete
toolchain with a front end to a language, a linker and a standard library.}

\section{After the A-2}

\todo{later developed the A-3 and AT-3/MATH-MATIC; FLOW-MATIC (B-0, 1956/58), \FTN, COBOL}

\begin{quotation}
B-0, as the business language compiler was designated, became available to 
UNIVAC customers at the start of 1958. Before its completion, Remington Rand 
merged with Sperry Gyroscope Corporation to form Sperry Rand. The marketing 
department of the new company renamed the business language FLOW-MATIC. (In 
addition, AT-3 was renamed MATH-MATIC.) The completed version of FLOW-MATIC had 
a rich library of oper-ational verbs that appeared to meet the application 
needs of most businesses.23 These verbs included editing commands so 
information could be formatted before output. Furthermore, FLOW-MATIC provided 
unparalleled flexibility in data designation, thus allowing file names to be 
given complicated descriptions.
\end{quotation}

\begin{quotation}
her superiors that the proposed business language would only be in English.  
B-0, as the business language compiler was designated, became available to 
UNIVAC customers at the start of 1958. Before its completion, Remington Rand 
merged with Sperry Gyroscope Corporation to form Sperry Rand. The marketing 
department of the new company renamed the business language FLOW-MATIC. (In 
addition, AT-3 was renamed MATH-MATIC.) The completed version of FLOW-MATIC had 
a rich library of oper-ational verbs that appeared to meet the application 
needs of most businesses.23 These verbs included editing commands so 
informa-tion could be formatted before output. Furthermore, FLOW-MATIC provided 
unparalleled fl exibility in data designation, thus allowing fi le names to be 
given complicated descriptions.
\end{quotation}

\begin{quotation}
This was what she originally called the data processing compiler in January, 
1955; it was soon to be known as "B-0," later as the "Procedure Translator" [KM 
57], and finally as flow-matic [HO 58, TA 60], This language used English 
words, somewhat as math-matic did but more so, and its operations concentrated 
on business applications. The following examples are typical of flow-matic 
operations;
\cite{history_of_computing_in_the_twentieth_century_1980}
\end{quotation}

\section{Laning and Zierler at MIT}

\todo{
Laning and Zierler, 1954. Early 1950s. Inspiration for Backus/\FTN.
Worked more like a modern compiler than Hopper's A-0 and A-1.
}
\begin{quotation}
The first programming system to operate in the sense of a modern compiler was 
developed by J. H. Laning and N. Zierler for the Whirlwind computer at the 
Massachusetts Institute of Technology in the early 1950s. They described their 
system, which never had a name, in an elegant and terse manual entitled "A 
Program for Translation of Mathematical Equations for Whirlwind I," distributed 
by MIT to about one-hundred locations in January 1954.26 It was, in John 
Backus's words, "an elegant concept elegantly realized." Unlike the UNIVAC 
compilers, this system worked much as modern compilers work; that is, it took 
as its input commands entered by a user, and generated as output fresh and 
novel machine code, which not only executed those commands but also kept track 
of storage locations, handled repetitive loops, and did other housekeeping 
chores. Laning and Zierler's "Algebraic System" took commands typed in familiar 
algebraic form and translated them into machine codes that Whirlwind could 
execute.27 (There was still some ambiguity as to the terminology: while Laning 
and Zierler used the word "translate" in the title of their manual, in the 
Abstract they call it an "interpretive program.")28
\cite{new-history-of-modern-computing}
\end{quotation}

\section{IBM Mathematical FORmula TRANslating System}

IBM did not feel that Aiken and the Harvard Computation Laboratory had
given them sufficient crediti for their contributions to the Mark I,
which Thomas Watson Sr. and the IBM folks bitter about the experience
and eager to produce their own device entirely in-house.
This device would become the Selective Sequence Electronic Calculator, or the SSEC.
It was built on 57th Street in Manhattan, and it was monstrous.
Roughly 50 by 100 feet with a giant console and hundreds of toggle switches and
tape units and relays behind glass panels; there were giant windows that allowed
passersby to see the machine in action.

One such passerby was John Backus, a recent Masters graduate from Columbia University.
He was intrigued by the machine, so he walked in and asked for a job.
Robert "Rex" Seeber gave him a puzzle, which he solved, and he was hired on the spot
\cite{backus_oral_history_2006}.

Before his studies, Backus was in the Army, stationed in Camp Stewart Georgia.

\todo{Backus wasn't really that involved in fortran II and III and IV.
The rest of the team took over. He never wrote much fortran and doesn't have many thoughts
about it today other that the "we should be making things higher level."}
\todo{Backus moved into functional programming. Was never fruitful.
He became a lab fellow at IBM after Watson Jr started the fellowship program,
so he could kinda play with functional programming and not really get much done.}
\todo{He didn't like Hopper's ideas for COBOL, thought they were crazy and too complex.
Didn't like algol either. Really just wanted to make things simpler and higher level
and easier to use. Pushed back against the priesthood.}

\begin{quotation}
Though the FORTRAN operator's manual was completed by the fall of 1956, the 
compiler itself was not distributed to IBM 704 installations until April 1957. 
Within a year after distribution, half of the IBM 704 installations were using 
FORTRAN to solve more than half of all mathematical prob-lems.16 Subsequently, 
compilers were produced for the IBM 705 and the IBM 650, quickly making FORTRAN 
the most widely used automatic program of its day. By 1961, UNIVAC users 
demanded a compatible FORTRAN compiler and aban-doned Hopper's MATH-MATIC.
\cite{grace_hopper_and_the_invention_of_the_information_age_2009}
\end{quotation}

\begin{quotation}
It describes the system which will be made available during late 1956, and is 
intended to permit planning and fortran coding in advance of that time [p. 1], 
Object programs produced by fortran will be nearly as efficient as those 
written by good programmers [p. 2], "Late 1956" was, of course, a euphemism for 
April 1957. Here is how Saul Rosen described fortran's debut [RO 64, p. 4]: Like 
most of the early hardware and software systems, fortran was late in delivery, 
and didn't really work when it was delivered.
\cite{history_of_computing_in_the_twentieth_century_1980}
\end{quotation}

\begin{quotation}
It is not my intention to give a complete description of either; hence this 
section will describe only the main highlights of FORTRAN development. The 
earliest significant document that seems to exist 1s one marked
"PRELIMINARY REPORT, Specifications for the IBM Mathematical FORmula 
TRANslating System, FORTRAN", dated November 10, 1954 and issued by the 
Programming Research Group, Applied Science Division, of IBM. 
\cite{sammet_programming_languages_history_and_fundamentals_1969}
\end{quotation}

\begin{quotation}
Laning and Zierler's algebraic compiler served as evidence that prestigious 
institutions such as MIT were taking automatic programming seriously, prompting 
Backus to write Laning a letter shortly after the May symposium. In the letter, 
Backus informed Laning that his team at IBM was working on a similar compiler, 
but that they had not yet done any programming or even any detailed planning.12 
To help formulate the specifi cations for their proposed language, Backus 
requested a demonstration of the algebraic compiler, which he and Ziller 
received in the summer of 1954. Much to their dismay, the two experienced fi 
rsthand the effi ciency dilemma of compiler-based language design. The MIT 
source code was commendable, but the compiler slowed down the Whirlwind 
computer by a factor of 10. Since computer time was so dear a commodity, Backus 
realized that only a com-piler that maximized effi ciency could hope to compete 
with human programmers. Despite this initial disappointment, Laning and 
Zierler's work inspired Backus to attempt to build a compiler that could 
translate a rich mathematical language into a suffi -ciently economical program 
at a relatively low cost.13
\cite{grace_hopper_and_the_invention_of_the_information_age_2009}
\end{quotation}

\pagebreak
\section{Timeline}
\input{chapters/dawn-timeline.tex}
